Edge sign prediction in general graphs and Correlation Clustering

### 2017-11-14 edge/model.tex
take feedback on edge chapter into account
- rewrite the setting part, giving the definition of s and g right away,
and adding a “prediction” operator
- try to motivate better the | x - b - \sum a⋅w | term in (1.2)
- acknowledge the signed case is only a sanity check (I couldn't find a
closed form solution for the noisy case)
- give a list of methods at the beginning of 1.2
- fix typos and math
- split the generation process in 4 steps and explicitly state its input
and output
- reorganize the results writeup in 4 parts: generation, measures,
initial AMI, commented tables

As in the previous chapter, we are given an undirected, unweighted graph $G=(V,E)$. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator $s$ between feature vectors, such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$,
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm.
While being exactly of norm one is not primordial, the role of bounding the norm of the directions
(\ie{} having $\sum_{i=1}^d {w_{uv}}_{;i}^2 \leq B^2$) is to tie the $d$ dimensions together.
Without such a constraint, we could simply study each dimension separately, whereas this bound
adds a coupling across different attributes, since increasing the weight of an attribute mechanically
affects the weights of the other attributes.

Depending on the application, and the semantic we ascribe to links between nodes, we can design and
interpret the combination operator $s$ as a difference, a similarity or a distance. Likewise, we
also have some freedom in defining how $s_{uv}$ and $w_{uv}$ are combined to assign a single real
value $g(s_{uv}, w_{uv})$ to the edge $(u,v)$. Let us call $g(w_{uv}, s_{uv})$ the \emph{goodness}
of $w_{uv}$ with respect to the edge $(u,v)$. We view this value as how adequately the direction
$w_{uv}$ \emph{explains}, or describes, the connection between $u$ and $v$.

\medskip

As written in the introduction, because we are interested in (partial) heterophilic links, we assume
in the following that both the profiles and the directions are vectors with components in $[-1, 1]$.
Let us take the example of encoding the gender of a user in a social network, assuming there is only
two values possible: male and female. Traditionally, this attribute would be set to $1$ if the user
is female and $0$ otherwise. Instead, we use $+1$ and $-1$, to clearly emphasize the difference
between the two values. The same is valid for continuous variables. For instance, an attribute could
range from $-1$ to $+1$ to represent the ideological stance of a user on the conservative/liberal
scale~\autocite{newsBias15}.\footnote{Note of course that in both cases, negative values do not
represent a judgement but are merely a reversible mathematical convenience.} This applies even to
non diverging positive quantities such as age, which can be projected to $[-1, 1]$ using
pre-processing such as standardization or $z$-score.

A natural choice for the operator $s$ is then the difference between the profiles, that is
$s_{uv} = x_u - x_v$. In that spirit, if $u$ and $v$ are close along a direction $w_{uv}$, we expect
${x_u}^T w_{uv}$ to be close to ${x_v}^T w_{uv}$, as measured by $|{x_u}^T w_{uv} - {x_v}^T
w_{uv}|$. Because we prefer to work with smooth functions, we define the goodness to be $g(s_{uv},
w_{uv}) = \left({s_{uv}}^T w_{uv} \right)^2$, and we want to minimize it with respect to $w_{uv}$.
When, on the \ith{} dimension, ${x_u}_{;i} - {x_v}_{;i}$ is close to zero, the minimization objective
and the fact that $w_{uv}$ is
unit-norm impose that ${w_{uv}}_{;i}$ should be large. On the other hand, with this formulation,
there is no way to set ${w_{uv}}_{;i}$ in order to \enquote{highlight} the fact that
${x_u}_{;i}$ and ${x_v}_{;i}$ are widely different.
There might be situations where this is not a problem. But again,
here we seek to model heterophilic interactions, at least along a subset of dimensions. Therefore,
we now introduce a more flexible approach.

Namely, we define $s$ to be the Hadamard (or component wise) product: $s_{uv} = x_u \circ x_v$, and
$g$ to be the dot product: $g(s_{uv}, w_{uv}) = {s_{uv}}^T w_{uv}$. The goodness of a direction $w$
for a given edge can thus be seen as a weighted sum of the per component similarity of $u$ and $v$
profiles. Therefore, it must be maximized with respect to $w_{uv}$. For a given component $i$, when
both ${x_u}_{;i}$ and ${x_v}_{;i}$ are large and of the same sign (that is close to either $+1$ or
$-1$), we see that $u$ and $v$ agree on the \ith{} dimension, and ${w_{uv}}_{;i}$ must accordingly
be a large positive value. On the other hand, when ${x_u}_{;i}$ and ${x_v}_{;i}$ are both large but
of different signs (\eg ${x_u}_{;i} = +1$ and ${x_v}_{;i} = -1$), $u$ and $v$ strongly disagree on
the \ith{} dimension, requiring in a large negative value for ${w_{uv}}_{;i}$.  Finally, if one of
${x_u}_{;i}$ or ${x_v}_{;i}$ is zero, denoting either an unimportant feature or a missing value,
then the value of ${w_{uv}}_{;i}$ does not matter, and will be set to zero by the maximization
procedure because of the unit norm constraint on $w_{uv}$.

\iffalse
\begin{aside}
When $s$ is a distance, and therefore positive, we impose $w_{uv}\geq 0$. A link $(u,v)$ is then
well explained by $w_{uv}$ when observed differences along this direction are small. More precisely,
when the quantity ${s_{uv}}^T w_{uv}$ is close to zero. Because $w_{uv}$ is unit norm, the \ith{}
component of $w_{uv}$ should be large when the \ith{} feature of $x_u$ and $x_v$ are close. Hence, a
null value for a component $i$ of $s_{uv}$ should not mean anything else than $u$ and $v$ share a
known and similar feature $i$ (\eg{} the unknown value should be different from $0$). When $s$ is
the difference $x_u - x_v$, it can be negative and the same reasoning can be applied, but with
$w_{uv}\in [-1,1]^d$ and a link value of $\left({s_{uv}}^T w_{uv}\right)^2$.

An alternative is to consider that the combination is a similarity. Links are created when the
observed similarity is high. Again, if $s$ is positive, it seems reasonable to impose that
$w_{uv}\geq0$ (and if $s$ can be negative, then $w_{uv}$ should be in $[-1,1]$). One main advantage
of similarities is maybe to avoid (or minimize) problems with missing values setting them to 0.
\end{aside}
\fi

\medskip

This second choice of $s$ and $g$ is indeed more flexible, for it encompasses all cases of
homophily, indifference and heterophily. Furthermore, since the profiles are given, the resulting
$s_{uv}$ can be seen as constants, meaning that $g$ is linear and therefore interpretable. Finally,
it also supports categorical data. We shall demonstrate all these features on a small example,
but we first define formally our problem.

@@

With working definitions of $s$ and $g$ at hand, we now formally write down the problem of
finding and assigning the most informative directions to every edge of $G$
\begin{problem}[\ecp{}]
  \label{p:edge_full}
  Given a graph $G=(V, E)$, node profiles $X\in [-1, 1]^{n\times d}$ and an integer $k \in \Nbb$,
  find a set of $k$ unit-norm directions $\mathcal{D}_k = \{w_1, w_2, \ldots, w_k\} \in \dsphere$
  and associate to every edge of $E$ the direction with the maximal goodness. Formally, solve
  \begin{equation}
    \label{eq:edge_full}
    \argmax_{w_1,\ldots,w_k \in \dsphere}
    \sum_{u,v \in E} \max_{\ell \in \rangesk} g(s_{uv}, w_{\ell})
  \end{equation}
  where $s_{uv} = x_u \circ x_v$ and $g(s_{uv}, w_{\ell}) = {s_{uv}}^T w_{\ell}$.
\end{problem}

@@

In \autoref{fig:edge_exe}, we present a small instance of the \ecp{} problem, where we are given the
graph shown in \autoref{fig:edge_exe_graph}, whose node attributes are listed
in Table~\ref{tab:edge_exe}. The first two attributes are categorical, denoting the company employing
each user, either \textsf{G} or \textsf{F}. Instead of a classic binary one-hot encoding, we
additionally use $-1$ to denote that a user had been employed by a given company. \textsf{Gender} is
encoded as presented earlier, and we also apply a pre processing to transform \textsf{age} between
$-1$ and $1$. Finally, \textsf{opinion} represents the view of former \textsf{F} employees about a
societal issue, with $0$ denoting a neutral position. In Table~\ref{tab:edge_exe}, we also show two
directions that explain the connections in the graph $G$. The first covers being of the same age and
same companies, especially at \textsf{G}, while the second directions covers former employees of
\textsf{F} having different \textsf{age} and \textsf{opinion}. Not only are the results
interpretable as claimed before, but we can also assess the strength of each explanation, using
their goodness. For instance, we see that the edge between $x$ and $y$ has a low score, because
although those two users have the same \textsf{age} and used to work at \textsf{F}, they do not
share a position at \textsf{G}.

\begin{figure*}[htpb]
  % \centering
  \begin{subfigure}[t]{0.55\textwidth}
    \centering
    \includegraphics[width=.9\textwidth]{tikz/edge_example_tikz.pdf}
    \caption{The graph $G$. Larger nodes represent old users while smaller ones are younger. Their
    gender is inscribed in the node and their shape represent their opinion: rectangle for $-1$,
    circle for $0$ and diamond for $+1$. Finally, orange stands for being employed at company G and
    faded blue for having been employed at company F. As for the edges, the red ones are labeled by
    the first direction and the green ones by the second direction, and we label them with the
    associated (normalized) goodness. } \label{fig:edge_exe_graph}
  \end{subfigure}~
  \begin{subfigure}[t]{0.42\textwidth}
    \centering
    \small
    \begin{tabular}{lrrrrr}
      \toprule
      {}    & G   & F    & gender & age    & opinion \\
      \midrule
      $u$   &  $+1$ &   $0$ &   $-1$ &  $-0.8$ &     $0$ \\
      $v$   &  $+1$ &  $-1$ &   $-1$ &  $-0.7$ &    $+1$ \\
      $w$   &  $+1$ &   $0$ &   $+1$ &   $0.6$ &    $+1$ \\
      $x$   &  $+1$ &  $-1$ &   $+1$ &   $0.9$ &    $-1$ \\
      $y$   &   $0$ &  $-1$ &   $-1$ &   $0.7$ &    $-1$ \\
      $z$   &   $0$ &  $-1$ &   $+1$ &  $-0.6$ &    $+1$ \\
      \midrule
      $w_1$ &  $+3$ &  $+1$ &    $0$ &    $+1$ &     $0$ \\
      $w_2$ &  $0$  &  $+4$ &    $0$ &    $-2$ &    $-4$ \\
      \bottomrule
    \end{tabular}
    \caption{The numerical attributes of the six nodes. They are followed by the (non normalized)
    two directions $w_1$ and $w_2$.} \label{tab:edge_exe}
  \end{subfigure}
  \caption{A small instance of \ecp{} and an handcrafted solution, albeit non-optimal.}
  \label{fig:edge_exe}
\end{figure*}

@@

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level.
For that we will consider node profiles to be normalized and having unit norm. A practical
justification is that if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. Furthermore, we assume that profiles
are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real weights and $b_u
\in \Rbb^d$ is the bias vector of $u$, \ie{} its inherent profile taken in isolation.
Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connects to another node $v$ along a direction $w_{uv}$, it \enquote{consumes} a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} =
  \sum_{u\in V} \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2_2
\end{equation}
from \eqref{eq:edge_full}.

\bigskip

We also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set $\forall i \in \rangesk\,, y_{uv,i} =
\frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} $, with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ term to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \left( \sum_{i = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} \right)
\end{equation}

These two constraints can be combined, so that the profiles are close to a linear combination of a
small number of directions.

@@

After having introduced \autoref{p:edge_full} and some additional constraints, we now present
methods to solve it. We start with a naive baseline, and propose a post-processing to improve it.
Then we describe a convex relaxation. Finally, we depart slightly from finding $k$ directions by
considering a low-rank matrix formulation, where each edge is assigned a linear combination of a
small number of base directions, paving the way to overlapping clustering.

@@

@@

Besides relaxing the $\max$ operator from \eqref{eq:edge_full}, another solution to remove it would
be to increase $k$ to be equal to $E$. As we already mentioned, this makes the problem trivial. To
avoid this, we add the constraint that all those $|E|$ directions are linear combinations of a small
number $k$ of basis directions. In order to formulate this more clearly, we rewrite the previous
objectives with matrices. We first show this can be expressed in a form that is amenable to the
Frank--Wolfe algorithm. Then we take the alternative road of explicit low-rank matrix factorization,
and we conclude by commenting on these approaches.

>>>>

Like we wrote in the introduction, to classify edges, we make use of both the topology of the graph
and the attributes of its nodes. Namely, we are given an undirected, unweighted graph $G=(V,E)$.  We
also have side information in addition to the graph topology. It takes the form, for every node $u$,
of a $d$-dimensional feature vector $x_u \in \Rbb^d$ that we call the \emph{profile} of $u$.
Because we want to model heterophily, we more explicitly have that $x_u \in [-1, 1]^d$. To justify
this statement, let us take the example of encoding the gender of a user in a social network,
assuming there is only two values possible: male and female. Traditionally, this attribute would be
set to $1$ if the user is female and $0$ otherwise. Instead, we use $+1$ and $-1$, to clearly
emphasize the difference between the two values. The same is valid for continuous variables. For
instance, an attribute could range from $-1$ to $+1$ to represent the ideological stance of a user
on the conservative/liberal scale~\autocite{newsBias15}.\footnote{Note of course that in both cases,
negative values do not represent a judgement but are merely a reversible mathematical convenience.}
This applies even to non diverging positive quantities such as age, which can be projected to $[-1,
1]$ using pre-processing like standardization or $z$-score.

Stacking all these node profiles in a $|V| \times d$ matrix $X$, our input so far is $(G, X)$. As it
is standard in classification problem, we are also given the number of edge types
$k$.\footnote{Because our problem is essentially unsupervised, assuming that we are given $k$ is a
simplification, on which we elaborate \vpageref{par:edge_choose_k}.} Moreover, recall that we do not
merely want to classify edges into type, but also provide an explanation of why two nodes are
connected. This requires the introduction of three further concepts:
\begin{inparaenum}[1)]
\item how to represent a connection (\ie{} an edge),
\item how to represent an edge's explanation, and
\item how to evaluate which explanation is the best.
\end{inparaenum}
We now present our modelling choice:

\begin{enumerate}[1),nosep,leftmargin=*]
  \item We naturally describe an edge $(u,v)$ through the profiles of its endpoints. More precisely,
    we combine $x_u$ and $x_v$ by an operator $s$ defined as:
    \begin{equation*}
      s(u, v) = x_u \circ x_v = s_{uv} \in \Rbb^d \, ,
    \end{equation*}
    where $\circ$ stands for the Hadamard (or component wise) product. As showed in
    \begin{marginfigure}
      % \small
      \captionsetup{font=small,type=table}
      \captionof{table}{We show the effect of the Hadamard product in combining information from two
        users $u$ and $v$. Instead of actual numbers, in this table we use $+$ to denote a large
        positive value, $-$ to denote a large negative value, $0$ to denote a small absolute value
        and $\star$ to denote an arbitrary value.
      \label{tab:edge_sim}}
      \begin{tabular}{ccc}
        \toprule
        ${x_u}_{;i}$ & ${x_v}_{;i}$ & ${s_{uv}}_{;i}$ \\
        \midrule
        $+$          & $+$          & $+$             \\
        $-$          & $-$          & $+$             \\
        $+$          & $-$          & $-$             \\
        $0$          & $\star$      & $0$             \\
        \bottomrule
      \end{tabular}
    \end{marginfigure}
    \autoref{tab:edge_sim}, for each dimension $i \in \rangesd$, the $s_{uv}$ holds a value
    indicating whether the profiles of $u$ and $v$ agree, disagree or are indifferent along that
    dimension.
  \item The $k$ explanations take the form of bounded norm vectors $\mathcal{D}_k = \{w_1, w_2,
    \ldots, w_k\} \subset \dball$, where $\dball$ is the $\ell_2$-unit ball of $\Rbb^d$. We call such a
    vector $w_\ell$, for $\ell \in \rangesk$, a \emph{direction}. As we shall explain more formally
    later on, those directions are exactly the parameters learned while solving our problem.
  \item Given an edge $(u,v)$, all the $k$ directions $\{w_1, w_2, \ldots, w_k\}$ provide a possible
    explanation of why $u$ and $v$ are connected. We score these explanations by assigning them a
    \emph{goodness} of explanation\footnote{Whereas the name is inspired by the concept of
    \emph{goodness of fit}, there is no formal relation.} defined as:
    \begin{equation*}
      g(s_{uv}, w_\ell) = {s_{uv}}^T w_\ell \, .
    \end{equation*}
    As the $w$ notation hints, a direction can be interpreted as a set of weights. These weights
    indicates to which extent each attribute contributes to the connection, according to that
    direction. For a given set of directions $\mathcal{D}_k$, the classification of edges is
    naturally performed by the following operator $\mathcal{E}$, returning the index of the
    direction with the maximal goodness:
    \begin{equation*}
      \begin{array}[t]{lrcl}
        \mathcal{E} : & E \times \left( \Rbb^d \times \ldots \times \Rbb^d \right)
                      & \longrightarrow & \rangesk \\
                      & (u,v), \mathcal{D}_k & \longmapsto
                      & \argmax_{\ell \in \rangesk} g(s_{uv}, w_\ell)
      \end{array}
    \end{equation*}
    Because it will be clear from context, we lighten the notation and simply write $\mathcal{E}(u,
    v)$ instead of $\mathcal{E}\left((u, v), \mathcal{D}_k \right)$.
\end{enumerate}

This choice of $s$ and $g$ enjoy two interesting properties. First, it encompasses all cases of
homophily, indifference and heterophily. For instance, given $x_u = \begin{pmatrix} 1 & 0 & -1
\end{pmatrix}^T$ and $x_v = \begin{pmatrix} -1 & 1 & -1 \end{pmatrix}^T$, the direction $w =
\begin{pmatrix} -\nicefrac{1}{\sqrt{2}} & 0 & \nicefrac{1}{\sqrt{2}} \end{pmatrix}^T \in \dball$
achieves the maximal goodness. In doing so, it highlights that $u$ and $v$ disagree on the first
component (\ie{} heterophily), agree on the last component (\ie{} homophily) and that the middle
component is irrelevant (\ie{} indifference). This also shows that, second, given the linear nature
of the goodness $g$ once $s_{uv}$ has been computed, it provides interpretable explanation of why
two nodes are connected.

Those two properties of $s$ and $g$ should not be taken for granted. Indeed, 
a natural choice for the operator $s$ is the difference between the profiles, that is
$s_{uv} = x_u - x_v$. In that spirit, if $u$ and $v$ are close along a direction $w_\ell$, we expect
${x_u}^T w_\ell$ to be close to ${x_v}^T w_\ell$, as measured by $|{x_u}^T w_\ell - {x_v}^T
w_\ell|$. And because it is easier to deal with smooth functions, we define the goodness to be $g(s_{uv},
w_\ell) = -\left({s_{uv}}^T w_\ell \right)^2$, where the minus preserve the semantic than the
\enquote{better} the direction $w_\ell$, the larger the goodness. However, with this formulation,
there is no way to set ${w_\ell}_{;i}$ in order to \enquote{highlight} the fact that ${x_u}_{;i}$
and ${x_v}_{;i}$ are widely different. Furthermore, the square in the goodness expression loses the
linearity.

We conclude with a last comment on the norm of the directions. While being exactly of norm one is
not primordial, the role of bounding the norm of the directions (\ie{} having $\sum_{i=1}^d
{w_\ell}_{;i}^2 \leq B^2$) is to tie the $d$ dimensions together. Without such a constraint, and we
our choice of $s$ and $g$, we could simply study each dimension separately, whereas this bound adds
a coupling across different attributes, since increasing the weight of an attribute mechanically
affects the weights of the other attributes.
For consistency, we also consider node profiles to be normalized and having unit norm. A further
practical justification is that if nodes are users of a social network and the profiles measure
their activity across several domains, this allows to compare users with various level of total
activity.

@@

With working definitions of $s$ and $g$ at hand, we now formally write down the problem of
finding the directions maximizing the goodness of the graph.
\begin{problem}[\ecp{}]
  \label{p:edge_full}
  Given a graph $G=(V, E)$, node profiles $X\in [-1, 1]^{n\times d}$ and an integer $k \in \Nbb$,
  find a set of $k$ norm-bounded directions $\mathcal{D}_k = \{w_1, w_2, \ldots, w_k\} \in \dball$
  and associate to every edge of $E$ the direction with the maximal goodness. Formally, solve
  \begin{equation}
    \label{eq:edge_full}
    \argmax_{\mathcal{D}_k = \{w_1,\ldots,w_k\} \subset \dball}
    \sum_{u,v \in E}  g(s_{uv}, w_{\mathcal{E}(u,v)})
  \end{equation}
  where $s_{uv} = x_u \circ x_v$, $g(s_{uv}, w_{\ell}) = {s_{uv}}^T w_{\ell}$ and $\mathcal{E}(u,v)
  = \argmax_{\ell \in \rangesk} g(s_{uv}, w_\ell)$.
\end{problem}

@@

In \autoref{fig:edge_exe}, we present a small instance of the \ecp{} problem, where we are given the
graph shown in \autoref{fig:edge_exe_graph}, whose node attributes are listed
in Table~\ref{tab:edge_exe}. The first two attributes are categorical, denoting the company employing
each user, either \textsf{G} or \textsf{F}. Instead of a classic binary one-hot encoding, we
additionally use $-1$ to denote that a user had been employed by a given company. \textsf{Gender} is
encoded as presented earlier, and we also apply a pre processing to transform \textsf{age} between
$-1$ and $1$. Finally, \textsf{opinion} represents the view of former \textsf{F} employees about a
societal issue, with $0$ denoting a neutral position. In Table~\ref{tab:edge_exe}, we also show two
directions that explain the connections in the graph $G$. The first covers being of the same age and
same companies, especially at \textsf{G}, while the second directions covers former employees of
\textsf{F} having different \textsf{age} and \textsf{opinion}. Not only are the results
interpretable as claimed before, but we can also assess the strength of each explanation, using
their goodness. For instance, we see that the edge between $x$ and $y$ has a low score, because
although those two users have the same \textsf{age} and used to work at \textsf{F}, not both of them
works at \textsf{G}.

\begin{figure*}[htpb]
  % \centering
  \begin{subfigure}[t]{0.55\textwidth}
    \centering
    \includegraphics[width=.9\textwidth]{tikz/edge_example_tikz.pdf}
    \caption{The graph $G$. Larger nodes represent old users while smaller ones are younger. Their
    gender is inscribed in the node and their shape represent their opinion: rectangle for $-1$,
    circle for $0$ and diamond for $+1$. Finally, orange stands for being employed at company G and
    faded blue for having been employed at company F. As for the edges, the red ones are labeled by
    the first direction and the green ones by the second direction, and we label them with the
    associated (normalized) goodness. } \label{fig:edge_exe_graph}
  \end{subfigure}~
  \begin{subfigure}[t]{0.42\textwidth}
    \centering
    \small
    \begin{tabular}{lrrrrr}
      \toprule
      {} & \textsf{G} & \textsf{F} & \textsf{gender} & \textsf{age} & \textsf{opinion} \\
      \midrule
      $u$   &  $+1$ &   $0$ &   $-1$ &  $-0.8$ &     $0$ \\
      $v$   &  $+1$ &  $-1$ &   $-1$ &  $-0.7$ &    $+1$ \\
      $w$   &  $+1$ &   $0$ &   $+1$ &   $0.6$ &    $+1$ \\
      $x$   &  $+1$ &  $-1$ &   $+1$ &   $0.9$ &    $-1$ \\
      $y$   &   $0$ &  $-1$ &   $-1$ &   $0.7$ &    $-1$ \\
      $z$   &   $0$ &  $-1$ &   $+1$ &  $-0.6$ &    $+1$ \\
      \midrule
      $w_1$ &  $+3$ &  $+1$ &    $0$ &    $+1$ &     $0$ \\
      $w_2$ &  $0$  &  $+4$ &    $0$ &    $-2$ &    $-4$ \\
      \bottomrule
    \end{tabular}
    \caption{The numerical attributes of the six nodes. All are categorical encoded with a ternary
      scheme, except for \textsf{age}, which has been transformed to fit in $[-1, 1]$.
      They are followed by the (non normalized) two directions $w_1$ and $w_2$.}
      \label{tab:edge_exe}
  \end{subfigure}
  \caption{A small instance of \ecp{} and an handcrafted solution, albeit non-optimal.}
  \label{fig:edge_exe}
\end{figure*}

@@

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level.
Indeed, one of our motivation is that it is not enough to look at edges in isolation, but rather we
expect the connections of one node to influence the connections of its neighbors. We now present two
of these local, node-level constraints.

First, if a graph was to perfectly follow our model, every edge $(u,v)$ would be maximally explained
by one of the $k$ directions. This would happen if the profiles of $u$ and $v$ were collinear with
this direction, up to sign reversal. When considering several edges, we could imagine that the nodes
profiles result from the following process. Two nodes $u$ and $v$ establish a connection by first
picking a direction $w_\ell \in \mathcal{D}_k$ and add to their base vector $b_u$ and $b_v \in
\Rbb^d$ a fraction $a_{uv} > 0$ of $w_\ell$. We therefore subtract the following term
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} =
  \sum_{u\in V} \left|\left| x_u - b_u -
  \sum_{v \in \nei(u)} a_{uv} w_{\mathcal{E}(u,v)} \right|\right|^2_2
\end{equation}
from \eqref{eq:edge_full}. The cost we pay is introducing the extra parameters $\{a_{uv}\}_{uv \in
E}$ and $\{b_u\}_{u \in V}$ to be learned, but we will see in the next section how to handle that in
practice. On the other hand, it has two advantages. First, this is another way, besides goodness, in
which the (given) profiles provide prior information about the (unknown) directions, and thus
restrain the search space. Second, this makes the directions dependent of the topology of the graph,
by coupling them with the neighborhood of each node.

Second, we also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set, $\forall i \in \rangesk$,
\begin{equation*}
  y_{uv,i} = \frac{\exp(\beta g(s_{uv}, w_i))}{\sum_{j=1}^k \exp(\beta g(s_{uv}, w_j))}\,,
\end{equation*}
with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ term to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \left( \sum_{i = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta g(s_{uv}, w_i))}{\sum_{j=1}^k \exp(\beta g(s_{uv}, w_j))} \right)
\end{equation}
It is reasonable to further expect that two neighboring nodes share one common direction among their
$k_\mathrm{local}$ ones. One way to ensure this condition is to imagine that the set of allowed
$k_\mathrm{local}$ directions changes smoothly over the graph. This property depends of the
topology, as we can see by considering two extreme cases: a line and a complete graph. On the line,
once a node has picked two allowed directions from its two neighbors, the remaining ones can be
chosen arbitrarily. On the complete graph on the other hand, this choice is must more constrained.

@@

We now introduce the five methods solving \autoref{p:edge_full} that we will experimentally compare
in the next section. Recall that given a graph topology $(V, E)$ and node profiles $X$, we first
build the edges representation $\mathcal{S} = \{s_{uv}\}_{(u,v) \in E}$ and then look for $k$
bounded directions that maximize the total goodness of the graph, possibly under topological
constraints. We start with a straightforward baseline, which simply performs a \newline
\hspace*{14pt} 1) \kmeans{} clustering of $\mathcal{S}$, \newline
and propose a post-processing to improve it, plugging our custom goodness metric into the existing
\newline \hspace*{14pt} 2) \lloyd{} algorithm for clustering. \newline
Because these two methods consider each edge in isolation with no regard for the topology, we then
describe a convex relaxation of the objective in equation \eqref{eq:edge_full}, which is further
\newline \hspace*{14pt} 3) \combined{} with the node constraint of equation
\eqref{eq:edge_node_loss}. \newline
Finally, we depart slightly from finding $k$ directions by
considering a low-rank matrix formulation, where each edge is assigned a linear combination of a
small number of base directions, paving the way to overlapping clustering. More precisely, we
give a convex formulation of that objective that can be solved by a \newline
\hspace*{14pt} 4) \fwa{} algorithm, \newline
as well an alternating optimization method, where we make the low-rank factorization \newline
\hspace*{14pt} 5) \pqt{}.

@@

In both cases, those methods do not take the topology of the graph into account, motivating the
following approaches.

@@

Besides relaxing the $\max$ operator from \eqref{eq:edge_full}, another solution to remove it would
be to increase $k$ to be equal to $|E|$. As we already mentioned, this makes the problem trivial. To
avoid this, we add the constraint that all those $|E|$ directions are linear combinations of a small
number $k$ of basis directions. In order to formulate this more clearly, we rewrite the previous
objectives with matrices. We first show this can be expressed in a form that is amenable to the
Frank--Wolfe algorithm. Then we take the alternative road of explicit low-rank matrix factorization,
and we conclude by commenting on these approaches.
